What Boolean Algebras Can Teach Us About Weighted Model Counting

### 2020-06-01 paper.tex
Experimental data

% TODO for later:
% 1) make up my mind about a,b vs x,y and stick to it (maybe x,y?)

@@

@@

% TODO: a special case for weight=0.
\begin{theorem}
  Let $\mathbf{B}$ be a free BA on $n$ generators $\{ l_i \}_{i=1}^n$ (for some
  $n \in \mathbb{N}$) with measure $m\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$,
  and let $L = \{ l_i \}_{i = 1}^n \cup \{ \neg l_i \}_{i = 1}^n$. Then there
  exists a weight function $w\colon L \to \mathbb{R}_{\ge 0}$ such that $m =
  \WMC_w$ if and only if
  \begin{equation} \label{eq:wmccondition}
  m(l \land l') = m(l)m(l')
  \end{equation}
  for all distinct $l, l' \in L$ such that $l \ne \neg l'$.
\end{theorem}

@@

A well-known way to overcome this limitation of independence is by adding more
literals \cite{DBLP:journals/ai/ChaviraD08}, i.e., extending the set $L$ covered
by the WMC weight function $w\colon L \to \mathbb{R}_{\ge 0}$. Let us translate this
idea to the language of Boolean algebras.

\begin{theorem} \label{thm:extension} % TODO: cite the fact about atoms
  Let $\mathbf{B}$ be a finite Boolean algebra freely generated by some set of
  `literals' $L$, and let $m\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$ be an
  arbitrary measure. We know that $\mathbf{B}$ has $n = 2^{|L|}$ atoms. Let
  $(a_i)_{i=1}^n$ denote those atoms in some arbitrary order. Let $L' = L \cup
  \{ \phi_i \mid i \in [n] \} \cup \{ \neg \phi_i \mid i \in [n] \}$ be the set
  $L$ extended with $2n$ new literals. Let $\mathbf{B'}$ be the unique Boolean
  algebra with
  \[
    \{ \phi_i \land a_i \mid i \in [n] \} \cup \{ \neg \phi_i \land a_i \mid i
    \in [n] \}
  \]
  as its set of atoms. Let $\iota\colon \mathbf{B} \to \mathbf{B'}$ be the inclusion
  homomorphism (i.e., $\iota(a) = a$ for all $a \in \mathbf{B}$). Let $w\colon L'
  \to \mathbb{R}_{\ge 0}$ be defined by
  \[
    w(l) = \begin{cases}
      \frac{m(a_i)}{2} & \text{if } l = \phi_i \text{ or } l = \neg\phi_i \text{
        for some } i \in [n] \\
      1 & \text{otherwise}
    \end{cases}
  \]
  for all $l \in L'$, and note that this defines a WMC measure $m'\colon \mathbf{B'}
  \to \mathbb{R}_{\ge 0}$. Then
  \[
    m(a) = (m' \circ \iota)(a)
  \]
  for all $a \in \mathbf{B}$.
\end{theorem}

In simpler terms, any measure can be computed using WMC by extending the BA with
more literals. More precisely, we are given the red part in
\[
  \begin{tikzcd}
    \textcolor{red}{\mathbb{R}_{\ge 0}} & & \\
    \textcolor{red}{\mathbf{B}} \arrow[red]{u}{m} \ar[r,hookrightarrow,"\iota"]
    & \mathbf{B'} \arrow{lu}[swap]{m'} & \\
    \textcolor{red}{L} \ar[u,red,hookrightarrow] \ar[r,hookrightarrow] & L'
    \ar[u,hookrightarrow] \arrow{r}{w} & \mathbb{R}_{\ge 0}
  \end{tikzcd}
\]
and construct the black part in such a way that the triangle commutes.

% TODO: make J depend on i
\begin{proof} % TODO: find a reference for this first claim
  Since $\mathbf{B}$ is freely generated by $L$, each atom $a_i \in \mathbf{B}$
  is an infimum of elements in $L$, i.e.,
  \[
    a_i = \bigwedge_{j \in J} a_{i,j}
  \]
  for some $\{ a_{i,j} \}_{j \in J} \subset L$. Moreover, each atom $b \in
  \mathbf{B'}$ can be represented as either $b = \phi_i \land a_i$ or $b =
  \neg\phi_i \land a_i$ for some atom $a_i \in \mathbf{B}$, also making it an
  infimum over a subset of $L'$. Then, for any $b \in \mathbf{B}$,
  \[
    (m' \circ \iota)(b) = \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\
        \phi_i \land a_i \le \iota(b)}} (w(\phi_i) + w(\neg\phi_i)) \prod_{j \in
    J} w(a_{i,j}),
  \]
  recognising that, for any $\iota(b)$, any atom $a_i \in \mathbf{B}$ satisfies
  $\phi_i \land a_i \le \iota(b)$ if and only if it satisfies $\neg\phi_i \land
  a_i \le \iota(b)$. Then, according to the definition of $w$,
  \[
    (m' \circ \iota)(b) = \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\
        \phi_i \land a_i \le \iota(b)}} (w(\phi_i) + w(\neg\phi_i)) =
    \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\ \phi_i \land a_i \le
        \iota(b)}} m(a_i) = m(b),
  \]
  provided that
  \[
    \phi_i \land a_i \le \iota(b) \quad \text{if and only if} \quad a_i \le b,
  \]
  but this is equivalent to
  \[
    \phi_i \land a_i = \phi_i \land a_i \land b \quad \text{if and only if}
    \quad a_i = a_i \land b
  \]
  which is true because $\phi_i \not\in L$.
\end{proof}

@@

\begin{conjecture}
  Let $\mathbf{B}$ and $\mathbf{B'}$ be Boolean algebras, and $\iota\colon
  \mathbf{B} \hookrightarrow \mathbf{B'}$ be the inclusion map such that
  $\mathbf{B}$ is freely generated by $L$, all atoms of $\mathbf{B'}$ can be
  expressed as meets of elements of $L'$, and the following subset relations are
  satisfied:
  \[
    \begin{tikzcd}
      \mathbf{B} \ar[r,hookrightarrow,"\iota"] & \mathbf{B'} \\
      L \arrow[Subset]{u}{} \arrow[Subset]{r}{} & L' \arrow[Subset]{u}{}
    \end{tikzcd}
  \]
  If, for any measure $m\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$, one can
  construct a weight function $w\colon L' \to \mathbb{R}_{\ge 0}$ such that the WMC
  measure $\WMC\colon \mathbf{B'} \to \mathbb{R}_{\ge 0}$ with respect to $w$
  satisfies
  \[
    m = \WMC \circ \iota,
  \]
  then $|L' \setminus L| \ge 2^{|L|+1}$.
\end{conjecture}
% \begin{proof}
%   % 1. An atom in B' must have more than just elements of L.
%   Let $a$ be an atom in $\mathbf{B}$, and let $b$ be an atom in $\mathbf{B'}$
%   such that $b \le a$. First, let us notice that as long as $|L| \ge
%   4$\footnote{Note that $|L|$ has to be an even number.}, $b \ne a$. Indeed, let
%   $p, r, \neg p, \neg r \in L$. Then
%   \begin{align*}
%     (\WMC \circ \iota)(p \land r) &= w(p)w(r), \\
%     (\WMC \circ \iota)(p \land \neg r) &= w(p)w(\neg r), \\
%     (\WMC \circ \iota)(\neg p \land r) &= w(\neg p)w(r), \\
%     (\WMC \circ \iota)(\neg p \land \neg r) &= w(\neg p)w(\neg r), \\
%   \end{align*}
%   But then we have that
%   \[
%     \frac{m(p \land r)}{m(\neg p \land r)} = \frac{w(p)}{w(\neg p)} =
%     \frac{m(p \land \neg r)}{m(\neg p \land \neg r)}.
%   \]
%   This places a condition on $m$, contradicting the assumption that the
%   construction works for an arbitrary $m$. Hence $b < a$.

@@

\subsection{Independence (all known results)}

@@

\subsection{Conditional Independence}

@@

\subsubsection{New Results}

\begin{theorem}
  Let $\{ \mathbf{A}_i \}_{i \in I}$ be a family of BAs with measures $\{
  m_i\colon \mathbf{A}_i \to \mathbb{R}_{\ge 0} \}_{i \in I}$, let $\mathbf{B}$
  be a BA, and let $\mathbf{A}$ be the amalgamated free product of $\{
  \mathbf{A}_i \}_{i \in I}$ over $\mathbf{B}$. Then there is a unique measure
  $m\colon \mathbf{A} \to \mathbb{R}_{\ge 0}$ such that, for any element $b \in
  \mathbf{B}$, finite subset $J \subseteq I$, and elements $\{ a_j \in
  \mathbf{A}_j \}_{j \in J}$,
  \[
    m \left(b \land \bigwedge_{j \in J} a_j \right) = \prod_{j \in J} m_j(b
    \land a_j).
  \]
\end{theorem}

@@

% TODO: make sure that the definition of 'free' is sufficient
\begin{theorem}[Pushouts of free BAs are free]
  Let
  \[
    \mathbf{A} = \bigoplus_{\substack{\mathbf{B}\\ i \in I}} \mathbf{A}_i
  \]
  be an amalgamated free product such that $\{ \mathbf{A}_{i} \}_{i \in I}$ are
  free BAs with $\{ S_i \}_{i \in I}$ as their respective sets of generators.
  Let $S = \bigcup_{i \in I} S_i$. Then $\mathbf{A}$ is a free BA with
  generating set $S$.
\end{theorem}
\begin{proof}
  Suppose we have a map from $S$ to an arbitrary BA $\mathbf{C}$, as in
  \[
    \begin{tikzcd}
      S_i \ar[r,hookrightarrow] \ar[d,hookrightarrow] & S \ar[d,hookrightarrow]
      \ar[ddr,bend left] & \\
      \mathbf{A}_i \ar[r,hookrightarrow] \ar[rrd,dashed,bend right] & \mathbf{A}
      \ar[rd,dashed]& \\
      & & \mathbf{C}.
    \end{tikzcd}
  \]
  We want to show that there exists a unique homomorphism $\mathbf{A} \to
  \mathbf{C}$. For all $i \in I$, from $S_i \hookrightarrow S$ and $S \to
  \mathbf{C}$ we get a map $S_i \to \mathbf{C}$, so---by the definition of a
  free BA---there is a unique homomorphism $\mathbf{A}_i \to \mathbf{C}$.
  Furthermore, a family of homomorphisms $\{ \mathbf{A}_i \to \mathbf{C} \}_{i
    \in I}$ uniquely determine a homomorphism $\mathbf{A} \to \mathbf{C}$ by the
  universal mapping property of a (wide) pushout. Thus $\mathbf{A}$ is a free BA
  with generating set $S$.
\end{proof}

>>>>

% Feedback from the panel: Clear up questions about independence between
% literals.

% TODO for later:
% 1) make up my mind about a,b vs x,y and stick to it (maybe x,y?)
% 2) Coproducts/pushouts vs (amalgamated) free products. Probably the former.
% 3) 'with generating set S' -> 'over S'

@@

\begin{definition}[\cite{koppelberg1989handbook}]
  Let $S$ be a set, and let $\mathbf{B}$ be a BA. Then $\mathbf{B}$ is a
  \emph{free BA over $S$} if there is a map $S \to \mathbf{B}$ such that for any
  BA $\mathbf{C}$ and map $S \to \mathbf{C}$, there is a unique homomorphism
  $\mathbf{B} \to \mathbf{C}$ that makes
  \[
    \begin{tikzcd}
      S \ar[rd] \ar[r] & \mathbf{B} \ar[d,dashed] \\
      & \mathbf{C}.
    \end{tikzcd}
  \]
  commute. A BA $\mathbf{B}$ is \emph{free} if $S$ exists.
\end{definition}

\begin{lemma}[\cite{sikorski1969boolean}] \label{lemma:finite_and_free}
  A finite BA is free if and only if it has $2^{2^n}$ elements for some $n \in
  \mathbb{N}$. It then has $2^n$ atoms and $n$ generators.
\end{lemma}

@@

% TODO: a special case for weight=0.
\begin{theorem}
  Let $\mathbf{B}$ be a free BA over $\{ l_i \}_{i=1}^n$ (for some $n \in
  \mathbb{N}$) with measure $m\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$, and let
  $L = \{ l_i \}_{i = 1}^n \cup \{ \neg l_i \}_{i = 1}^n$. Then there exists a
  weight function $w\colon L \to \mathbb{R}_{\ge 0}$ such that $m = \WMC_w$ if
  and only if
  \begin{equation} \label{eq:wmccondition}
  m(l \land l') = m(l)m(l')
  \end{equation}
  for all distinct $l, l' \in L$ such that $l \ne \neg l'$.
\end{theorem}

@@

Given this requirement for independence, a well-known way to represent
probability distributions that do not consist entirely of independent variables
is by adding more literals \cite{DBLP:journals/ai/ChaviraD08}, i.e., extending
the set $L$ covered by the WMC weight function $w\colon L \to \mathbb{R}_{\ge
  0}$. Let us translate this idea to the language of BAs.

% TODO: If L denotes literals, then it doesn't denote a generating set
\begin{theorem} \label{thm:extension}
  Let $\mathbf{B}$ be a free BA over a finite set $S$, and let $m\colon
  \mathbf{B} \to \mathbb{R}_{\ge 0}$ be an arbitrary measure. Let $L = \{ s \mid
  s \in S \} \cup \{ \neg s \mid s \in S \}$. By \cref{lemma:finite_and_free},
  we know that $\mathbf{B}$ has $n = 2^{|S|}$ atoms. Let $\{a_i\}_{i=1}^n$
  denote those atoms in some arbitrary order. Let $L' = L \cup \{ \phi_i
  \}_{i=1}^n \cup \{\neg \phi_i \}_{i=1}^n$ be the set $L$ extended with $2n$
  new elements. Let $\mathbf{B'}$ be the unique Boolean algebra with $\{ \phi_i
  \land a_i \}_{i=1}^n \cup \{ \neg \phi_i \land a_i \}_{i=1}^n$ as its set of
  atoms. Let $\iota\colon \mathbf{B} \hookrightarrow \mathbf{B'}$ be the
  inclusion homomorphism. Let $w\colon L' \to \mathbb{R}_{\ge 0}$ be defined by
  \[
    w(l) = \begin{cases}
      m(a_i)/2 & \text{if } l = \phi_i \text{ or } l = \neg\phi_i \text{ for
        some } i \in [n] \\
      1 & \text{otherwise}
    \end{cases}
  \]
  for all $l \in L'$, and note that this defines a WMC measure $m' =
  \WMC_{w}\colon \mathbf{B'} \to \mathbb{R}_{\ge 0}$. Then
  \[
    m(a) = (m' \circ \iota)(a)
  \]
  for all $a \in \mathbf{B}$.
\end{theorem}

In other words, any measure can be computed using WMC by extending the BA with
more literals. More precisely, we are given the left-hand column in
\[
  \begin{tikzcd}
    \textcolor{red}{\mathbb{R}_{\ge 0}} & & \\
    \textcolor{red}{\mathbf{B}} \arrow[red]{u}{m} \ar[r,hookrightarrow,"\iota"]
    & \mathbf{B'} \arrow{lu}[swap]{m'} & \\
    \textcolor{red}{L} \ar[u,red,hookrightarrow] \ar[r,hookrightarrow] & L'
    \ar[u,hookrightarrow] \arrow{r}{w} & \mathbb{R}_{\ge 0}
  \end{tikzcd}
\]
and construct the remaining part in such a way that the triangle commutes.

% TODO: make J depend on i
\begin{proof} % TODO: find a reference for this first claim
  Since $\mathbf{B}$ is free over $S$, each atom $a_i \in \mathbf{B}$ is an
  infimum of elements in $L$, i.e.,
  \[
    a_i = \bigwedge_{j \in J} a_{i,j}
  \]
  for some $\{ a_{i,j} \}_{j \in J} \subset L$. Moreover, each atom $b \in
  \mathbf{B'}$ can be represented as either $b = \phi_i \land a_i$ or $b =
  \neg\phi_i \land a_i$ for some atom $a_i \in \mathbf{B}$, also making it an
  infimum over a subset of $L'$. Then, for any $b \in \mathbf{B}$,
  \[
    (m' \circ \iota)(b) = \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\
        \phi_i \land a_i \le \iota(b)}} (w(\phi_i) + w(\neg\phi_i)) \prod_{j \in
    J} w(a_{i,j}),
  \]
  recognising that, for any $\iota(b)$, any atom $a_i \in \mathbf{B}$ satisfies
  $\phi_i \land a_i \le \iota(b)$ if and only if it satisfies $\neg\phi_i \land
  a_i \le \iota(b)$. Then, according to the definition of $w$,
  \[
    (m' \circ \iota)(b) = \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\
        \phi_i \land a_i \le \iota(b)}} (w(\phi_i) + w(\neg\phi_i)) =
    \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\ \phi_i \land a_i \le
        \iota(b)}} m(a_i) = m(b),
  \]
  provided that
  \[
    \phi_i \land a_i \le \iota(b) \quad \text{if and only if} \quad a_i \le b,
  \]
  but this is equivalent to
  \[
    \phi_i \land a_i = \phi_i \land a_i \land b \quad \text{if and only if}
    \quad a_i = a_i \land b
  \]
  which is true because $\phi_i \not\in L$.
\end{proof}

@@

\begin{conjecture}
  Let $\mathbf{B}$ and $\mathbf{B'}$ be Boolean algebras, and $\iota\colon
  \mathbf{B} \hookrightarrow \mathbf{B'}$ be the inclusion map such that
  $\mathbf{B}$ is free over $L$, all atoms of $\mathbf{B'}$ can be
  expressed as meets of elements of $L'$, and the following subset relations are
  satisfied:
  \[
    \begin{tikzcd}
      \mathbf{B} \ar[r,hookrightarrow,"\iota"] & \mathbf{B'} \\
      L \arrow[Subset]{u}{} \arrow[Subset]{r}{} & L' \arrow[Subset]{u}{}
    \end{tikzcd}
  \]
  If, for any measure $m\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$, one can
  construct a weight function $w\colon L' \to \mathbb{R}_{\ge 0}$ such that the WMC
  measure $\WMC\colon \mathbf{B'} \to \mathbb{R}_{\ge 0}$ with respect to $w$
  satisfies
  \[
    m = \WMC \circ \iota,
  \]
  then $|L' \setminus L| \ge 2^{|L|+1}$.
\end{conjecture}
% \begin{proof}
%   % 1. An atom in B' must have more than just elements of L.
%   Let $a$ be an atom in $\mathbf{B}$, and let $b$ be an atom in $\mathbf{B'}$
%   such that $b \le a$. First, let us notice that as long as $|L| \ge
%   4$\footnote{Note that $|L|$ has to be an even number.}, $b \ne a$. Indeed, let
%   $p, r, \neg p, \neg r \in L$. Then
%   \begin{align*}
%     (\WMC \circ \iota)(p \land r) &= w(p)w(r), \\
%     (\WMC \circ \iota)(p \land \neg r) &= w(p)w(\neg r), \\
%     (\WMC \circ \iota)(\neg p \land r) &= w(\neg p)w(r), \\
%     (\WMC \circ \iota)(\neg p \land \neg r) &= w(\neg p)w(\neg r), \\
%   \end{align*}
%   But then we have that
%   \[
%     \frac{m(p \land r)}{m(\neg p \land r)} = \frac{w(p)}{w(\neg p)} =
%     \frac{m(p \land \neg r)}{m(\neg p \land \neg r)}.
%   \]
%   This places a condition on $m$, contradicting the assumption that the
%   construction works for an arbitrary $m$. Hence $b < a$.

@@

\subsection{Preliminaries}

@@

@@

\subsection{New Results}

% TODO: make sure that this is enough to guarantee a pushout
\begin{theorem}
  Let $\{ S_i \}_{i=0}^n$ be a finite set of finite sets for some $n > 1$ such
  that for all distinct positive integers $i$ and $j$, $S_i \cap S_j = S_0$, and
  let
  \[
    \mathbf{A} = \bigoplus_{\substack{\mathcal{F}(S_0)\\ 1 \le i \le n}}
    \mathcal{F}(S_i).
  \]
  Let $(m_i\colon \mathcal{F}(S_i) \to \mathbb{R}_{\ge 0})_{i=1}^n$ be arbitrary
  measures. Then there is a unique measure $m\colon \mathbf{A} \to
  \mathbb{R}_{\ge 0}$ such that, for any element $b \in \mathcal{F}(S_0)$,
  subset $J \subseteq \{ 1, 2, \dots, n \}$, and elements $\{ a_j \in
  \mathcal{F}(S_j \setminus S_0) \}_{j \in J}$,
  \[
    m \left(b \land \bigwedge_{j \in J} a_j \right) = \prod_{j \in J} m_j(b
    \land a_j).
  \]
\end{theorem}
\begin{proof}
  % TODO: start here
\end{proof}

@@

\begin{theorem}[Pushouts of free BAs are free]
  Let
  \[
    \mathbf{A} = \bigoplus_{\substack{\mathbf{B}\\ i \in I}} \mathbf{A}_i
  \]
  be an amalgamated free product such that $\{ \mathbf{A}_{i} \}_{i \in I}$ are
  free BAs with $\{ S_i \}_{i \in I}$ as their respective sets of generators.
  Let $S = \bigcup_{i \in I} S_i$. Then $\mathbf{A}$ is a free BA with
  generating set $S$.
\end{theorem}
\begin{proof}
  Suppose we have a map from $S$ to an arbitrary BA $\mathbf{C}$, as in
  \[
    \begin{tikzcd}
      S_i \ar[r,hookrightarrow] \ar[d,hookrightarrow] & S \ar[d,hookrightarrow]
      \ar[ddr,bend left] & \\
      \mathbf{A}_i \ar[r,hookrightarrow] \ar[rrd,dashed,bend right] & \mathbf{A}
      \ar[rd,dashed]& \\
      & & \mathbf{C}.
    \end{tikzcd}
  \]
  We want to show that there exists a unique homomorphism $\mathbf{A} \to
  \mathbf{C}$. For all $i \in I$, from $S_i \hookrightarrow S$ and $S \to
  \mathbf{C}$ we get a map $S_i \to \mathbf{C}$, so---by the definition of a
  free BA---there is a unique homomorphism $\mathbf{A}_i \to \mathbf{C}$.
  Furthermore, a family of homomorphisms $\{ \mathbf{A}_i \to \mathbf{C} \}_{i
    \in I}$ uniquely determine a homomorphism $\mathbf{A} \to \mathbf{C}$ by the
  universal mapping property of a (wide) pushout. Thus $\mathbf{A}$ is a free BA
  with generating set $S$.
\end{proof}
